TacticToe: Learning to Prove with Tactics

### 2018-02-19 tactictoe_jv.tex
Sec 6

The exploration term is determined by the prior policy and the current policy.
The current policy is calculate from the number of times a node $x$ has been
traversed during node selection, noted $\mathit{Visit}(x)$.

@@

We now concentrate on the definition of a reasonable evaluation function for the
output of tactics. Theoretically, a list of goals is interesting if short proofs
are possible each the goals in the list, and the lengths of the proofs can be
estimated from previous proof searches. However, this
approach generates too much data and slows down the proof search. So, we only
collect outputs of tactics tested during orthogonalization. We
declare a list of goals $l$ to be positive if it is has been produced by the
winner of an orthogonalization competition.
The set of positive examples in
$\mathit{Predictions}^{\mathit{goal\_list}}_k(l)$ is noted
$\mathit{Positive}^{\mathit{goal\_list}}_k(l)$.
We chose $k=10$ as a default evaluation radius in our experiments.
And we evaluate a node $a$ through the list of goals $G(a)$ it
contains using the prior evaluation function:
\begin{align*}
\mathit{PriorEvaluation}_k (a) &=
  \frac{card\ \mathit{Positive}^{\mathit{goal\_list}}_k(G(a)}{k}\\
\end{align*}

@@

\section{Orthogonalization}\label{sec:ortho}
Different tactics may transform a single goal in the same way. Exploring such
equivalent paths
is undesirable, as it leads to inefficiency in automated proof search.
To solve this problem, we modify how the database of tactics is build.
Each time a new tactic-goal pair $(t,g)$ is extracted from a tactic proof and
about to be recorded, we consider if there does not exists already a better
tactic for $g$ in our database. To this end, we organize a competition between
the
$k$ closest tactic-goal pairs $\mathit{Predictions}^{\mathit{tactic}}_k(g)$.
In our experiments,
the default orthogonalization radius is $k=20$.
The winner is the tactic that subsumes (see Definition~\ref{def:tacsub}) the
original tactic on $g$ and that
appears in the largest number of tactic-goal pairs in the database.
The winning tactic $w$ is then associated with $g$, producing the pair $(w,g)$
and is stored in the database instead of the original pair $(t,g)$.
As a result, already successful tactics with a large coverage are preferred,
and new tactics are considered only if they provide a different contribution.
We now give a formal definition of the concepts of subsumption and coverage
that are required for expressing the orthogonalization algorithm.

@@

\begin{definition} (Goal subsumption)\\
A goal subsumption $\le$ is a partial relation on a set of goals.\\
Assuming we have a way to estimate the number steps required to solve a goal,
a possibly useful subsumption definition could be:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ has a shorter proof than }
g_2\]
By default, we choose a minimal subsumption defined by:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent
to } g_2\]
\end{definition}

@@

This allows us to define subsumption for tactics.
\begin{definition}\label{def:tacsub}(Tactic subsumption)\\
Given two non-failing tactics $t_1$ and $t_2$ on $g$, a tactic $t_1$ subsumes a
tactic $t_2$ on a goal $g$, noted $\le_g$, when:

@@

In the end, the winning tactic of the orthogonalization competition can be
expressed by the formula:

@@

In Example~\ref{ex:cs3}, we prove a theorem about lists, a domain where \tactictoe excels compared to ATPs.
Given two list $l_1$ and $l_2$ where the length of $l_1$ (noted $p$) is less
than a natural
number $n$, the theorem states updating the $n^{th}$
element of the concatenation of $l_1$ and $l_2$ is the same as
updating the $m^{th}$ element of $l_2$ where $m = n - p$. Again, the
\tactictoe's proof is
much more readable. It starts by applying induction on $l$. It solves the base
case of the induction by rewriting and proceeds by cases on $n=0$ or $n>0$
in the induction hypothesis. It finalizes the proof with a short call to \metis
using the definition of
\texttt{LUPDATE}. Here, \tactictoe's proof is arguably smaller, faster and
easier to
understand and maintain. Such proofs after an expert review could replace the
their respective original human proofs in the \holfour repository.

>>>>

The exploration term is determined by the prior policy and the current policy.
The current policy is calculate from the number of times a node $x$ has been
traversed during node selection, denoted $\mathit{Visit}(x)$.

@@

We now concentrate on the definition of a reasonable evaluation function for the
output of tactics. Theoretically, a list of goals is interesting if short proofs
are possible each the goals in the list, and the lengths of the proofs can be
estimated from previous proof searches. However, this
approach generates too much data and slows down the proof search. So, we only
collect outputs of tactics tested during orthogonalization. We
declare a list of goals $l$ to be positive if it is has been produced by the
winner of an orthogonalization competition.
The set of positive examples in
$\mathit{Predictions}^{\mathit{goal\_list}}_k(l)$ is denoted
$\mathit{Positive}^{\mathit{goal\_list}}_k(l)$.
We chose $k=10$ as a default evaluation radius in our experiments.
And we evaluate a node $a$ through the list of goals $G(a)$ it
contains using the prior evaluation function:
\begin{align*}
\mathit{PriorEvaluation}_k (a) &=
  \frac{card\ \mathit{Positive}^{\mathit{goal\_list}}_k(G(a)}{k}\\
\end{align*}

@@

\section{Orthogonalization}\label{sec:ortho}
Different tactics may transform a single goal in the same way. Exploring such
equivalent paths
is undesirable, as it leads to inefficiency in automated proof search.
To solve this problem, we modify the construction of the tactics database.
Each time a new tactic-goal pair $(t,g)$ is extracted from a tactic proof and
about to be recorded, we consider if there does not already exist a better
tactic for $g$ in our database. To this end, we organize a competition between
the
$k$ closest tactic-goal pairs $\mathit{Predictions}^{\mathit{tactic}}_k(g)$.
In our experiments,
the default orthogonalization radius is $k=20$.
The winner is the tactic that subsumes (see Definition~\ref{def:tacsub}) the
original tactic on $g$ and that
appears in the largest number of tactic-goal pairs in the database.
The winning tactic $w$ is then associated with $g$, producing the pair $(w,g)$
and is stored in the database instead of the original pair $(t,g)$.
As a result, already successful tactics with a large coverage are preferred,
and new tactics are considered only if they provide a different contribution.
We now give a formal definition of the concepts of subsumption and coverage
that are required for expressing the orthogonalization algorithm.

@@

\begin{definition} (Goal subsumption)\\
A goal subsumption $\le$ is a partial relation goals.
Assuming we have a way to estimate the number steps required to solve a goal,
a possibly useful subsumption definition could be: \todo{shorter or equal}
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ has a shorter proof than }
g_2\]
By default, we choose a minimal subsumption defined by:
\[g_1 \le g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent
to } g_2\]
\end{definition}
\todoi{@TG: Which of the two above?}

@@

This allows us to define subsumption for tactics.
\begin{definition}\label{def:tacsub}(Tactic subsumption)\\
Given two non-failing tactics $t_1$ and $t_2$ on $g$, a tactic $t_1$ subsumes a
tactic $t_2$ on a goal $g$, denoted $\le_g$, when:

@@

Finally, the winning tactic of the orthogonalization competition can be
expressed by the formula:

@@

In Example~\ref{ex:cs3}, we prove a theorem about lists, a domain where \tactictoe excels compared to ATPs.
Given two list $l_1$ and $l_2$ where the length of $l_1$ (denoted $p$) is less
than a natural
number $n$, the theorem states updating the $n^{th}$
element of the concatenation of $l_1$ and $l_2$ is the same as
updating the $m^{th}$ element of $l_2$ where $m = n - p$. Again, the
\tactictoe's proof is
much more readable. It starts by applying induction on $l$. It solves the base
case of the induction by rewriting and proceeds by cases on $n=0$ or $n>0$
in the induction hypothesis. It finalizes the proof with a short call to \metis
using the definition of
\texttt{LUPDATE}. Here, \tactictoe's proof is arguably smaller, faster and
easier to
understand and maintain. Such proofs after an expert review could replace the
their respective original human proofs in the \holfour repository.
